Characterizing subarctic biomes for land surface modeling of pollution and climate risk

### 2021-05-10 VegOzone.tex
Refine section 2 and 3

According to a report by the Norwegian Meteorological institute \citep{MetNOR2019}, the summer of 2018 was the warmest and driest ever recorded in eastern, western, and southern Norway. In the north (including Finnmark), it was amongst the warmest on record -- favorable conditions for ground-level ozone formation.
An unusually weak and northward shifted jet stream allowed for a persistent high-pressure system above northern Europe, including Fennoscandia which blocked the low-pressure systems for several consecutive months. In the period May -- July, southern Norway had temperatures $4\,\unit{^\circ C}$ above normal. While southern Norway had only about $60\,\unit{\%}$ of normal precipitation, northern Norway as a whole had close to normal precipitation, but with local variations.
Thermal stress on vegetation was exceptional not only in large parts of Fennoscandia but also in much of Europe, where the influence of the high-pressure system extended even over five months (April/May, July--September). These conditions gave rise to massive forest fires in different parts of Europe and thus an increase in ozone precursors. Boreal wildfires emit in addition to \chem{CO_2} also \chem{CO} ($\chem{[CO]}/\chem{[CO_2]} \propto 6-13\,\unit{\%}$) and VOCs ($\chem{[VOCs]}/\chem{[CO_2]} \propto 0.5-1.5\,\unit{\%}$) \citep{AE:Cofer1990}.
A total of 2079 forest fires of different sizes were registered in Norway in 2018, twice as many as in the preceding years 2016/17 \citep{DSB2019}, last accessed April 2020). In Sweden, about 500 fires had been reported (five times more than in a usual summer), and an estimated total of $25000\,\unit{hectare}$ burned down in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) \citep{SOU2019}. Coincident peak \chem{[O_3]} is found in ozone monitoring data from Svanhovd but also over whole northern Fennoscandia in July (see Fig.~\ref{fig:data_svanvik_2018_2019} or Fig.~\ref{fig:ozone_reconstruction_2018_07}).
During the 2003 drought period, elevated ozone in Europe was promoted by a combination of various factors such as wildfires, reduced cloud cover (increased solar radiation), reduced dry deposition and turbulent mixing due to the stagnant weather conditions, and increased BVOC emissions \citep{JGR:Solberg2018}. It is likely that these factors also contributed to enhanced \chem{[O_3]} in 2018.\\

@@

The observation site at NIBIO Environment Centre Svanhovd and all related data are presented in Section~\ref{sec:data}. In Section~\ref{sec:stats}, we derive multi-annual means for ground-level ozone concentration and key environmental variables (temperature, precipitation, and global irradiance) based on in situ observations and evaluate the deviation of 2018/19 from the norm. We use the $\mathrm{DO_3SE}$ model to estimate the ozone uptake by natural and semi-natural vegetation to investigate the main environmental drivers for projected ozone damage risk in 2018 (Section~\ref{sec:do3se}). In this end, we derive bespoke parameterizations for subarctic species to assess systematic uncertainties arising from the use of default generic parameterizations established from less climatically extreme conditions. In Section~\ref{sec:conc}, we will summarize our results and discuss how this research can support improvements in future assessments of pollution risk and more generally bio-biogeochemical cycling.

@@

Hourly averaged $2\,\unit{m}$ temperatures above $20\,\unit{^\circ C}$ occurred more regularly in July 2018 than in 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}b). In 2018, spring temperature regularly rose above freezing only in May, while in 2019 this occurred already early in March/April.
More rain events with accumulated daily precipitation ($\sum_d \mathrm{Precip}$) above $10\,\unit{mm}$ occurred in the summer of 2018 compared to 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}c).
Qualitatively, global irradiance ($Q_0$) displayed in Fig.~\ref{fig:data_svanvik_2018_2019}d) was higher in May and July 2018 compared to 2019, while June 2019 showed higher irradiance than 2018. A high global irradiance in most cases is the result of a low cloud fraction. In both years, the maximum recorded $Q_0$ was roughly $750\,\unit{W\,m^{-2}}$ reached in June.

@@

A comprehensive overview over all past and present ozone observation sites in northern Fennoscandia is given in Table~\ref{tab:ebas_obs} and Fig.~\ref{fig:station_map_fennoscandia}, respectively. All stations with available long-term observations are located at higher altitude than Svanhovd. In the absence of local precursors, the stratification of the atmosphere usually leads to a higher abundances in ozone at higher altitude \citep[e.g.,][]{AB:Klingberg2009}. The periphery vegetation at all sites is similar and consists mainly of pine forests, birch, and heath. However, on closer inspection the vegetation differs substantially. Ozone measurements at Pallas are taken on top of mount Sammaltunturi $100\,\unit{m}$ above treeline (pine forest) and the surrounding vegetation mainly consists of low vascular plants, mosses, and lichen \citep{BER:Hatakka2003}. At Svanhovd, the ozone measurements are taken amidst a meadow surrounded by agriculturally managed herbaceous plants and grass adhering to the \citet{WMOGuide2018} standards (Fig.\ref{fig:svanhovd_research_station}). In conjunction with the location at lower altitude, this results in generally lower \chem{[O_3]} at Svanhovd compared to Pallas.

Data from before the 1990s are available solely from Svanvik and Jergul. These data, however, do not follow the high-quality standard procedures implemented nowadays and have to be treated with care \citep{NILU2003}. Especially, ozone monitors did not undergo regular re-calibration. This probably leads to drifts in the observed data and may impose a false periodicity. In 1997, the ozone monitoring at Jergul was relocated downstream of the river Karasjokka to a site close by Karasjok. Although the same equipment has been used, this might have introduced a baseline shift in the combined data series, denoted as Jergul/Karasjok, which would be particularly problematic if these were the sole data for comparison. Only a few data have been taken upstream the Pasvik river at Janiskovski, hence these will not be taken into further consideration, but are listed for completeness. There is an overlap of about $4\,\unit{years}$ between data taken at Esrange, Jergul/Karasjok, and Svanvik (Supplement~Fig.~\ref{fig:ozone_timesseries_fenoscandic_obs}).

@@

\subsubsection{Temperature, precipitation, and global irradiance}
\label{subsubsec:clim_temp_prec}
The location of Svanvik at $69.45\,\unit{^\circ N}$ suggests a subarctic climate. The derived climate diagram based on climatological data from the atmospheric monitoring in Svanvik (1992--2012) (Fig.~\ref{fig:climatediagram}) supports this very well. Monthly averaged $2\,\unit{m}$ temperatures (red line) are displayed with a standard error of the mean (SE) as error band. We chose SE in this case due to the large interannual variability in temperature reflected in the standard deviation. Averaged monthly accumulated precipitation (blue bars) is shown with standard deviation as error bars. Months with average temperatures below freezing are denoted with a star. As can be seen, temperatures stay below freezing for 5 consecutive months, while only 2 months breach $10\,\unit{^\circ C}$ regularly (July, August), satisfying the conditions for K\"{o}ppen's climate classification of a regular subarctic climate (Dfc) \citep[][e.g.]{SD:Beck2018}. The highest monthly average temperature is $(13.1\pm 1.1)\,\unit{^\circ C}$ in July and the lowest $(-12.8\pm 2.0)\,\unit{^\circ C}$ in February. In the period 1992--2012, the coldest measured temperature was $-45.2\,\unit{^\circ C}$ observed January 27, 1999, while the highest temperature ($29.4\,\unit{^\circ C}$) occurred July 16 the same year.

@@

The derived climatology for global irradiance ($Q_0$) at Svanvik (Appendix Fig.~\ref{fig:global_rad_clim}a)) approaches zero in December/January (polar night) and reaches its maximum in June/July (midnight sun conditions). The highest observed averaged daily maximum ($\left<Q_0^\mathrm{max}\right>$) amounts to $800\,\unit{W\,m^{-2}}$, while the averaged daily mean ($\left<Q_0^\mathrm{mean}\right>$) is $450\,\unit{W\,m^{-2}}$ (regularly reached in May--July). The maximum of the averaged daily minimum ($\left<Q_0^\mathrm{min}\right>$) lies at about $200\,\unit{W\,m^{-2}}$.

\subsubsection{Ozone}
\label{subsubsec:clim_ozone}
In the following, we derive an ozone climatology (\chem{\left<[O_3]\right>}) for Svanvik and for northern Fennoscandia. To this end, we first calculate a pairwise correlation between the different stations. The correlation of \chem{[O_3]} between Esrange, Pallas, and Jergul/Karasjok is fairly high ($r^2\approx 0.8$, Appendix Fig.~\ref{fig:density_distribution}a, b). Therefore, we select these data to compute the generalized ozone climatology for northern Fennoscandia. The correlation between Svanvik with any of the active monitoring sites in 2018 (Esrange and Pallas) is rather low, 0.4 and 0.6, respectively. Thus, conditions at Svanvik cannot directly be represented by measurements taken at either Esrange or Pallas.

In Figure~\ref{fig:ozone_climatology_fenoscandic_obs_spline}, two dimensional density distributions of \chem{[O_3]} for the combined northern Fennoscandia data (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}a) and Svanvik (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) are shown. Darker colors indicate higher probability to observe these values. On top of the density distributions, a $10\,\unit{days}$ average of daily mean (\chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}}) is displayed together with $1 \sigma$ uncertainties and SE, respectively. Splines have been fitted through the data for easier inter-comparison.

The generalized northern Fennoscandic ozone climatology peaks at $48\,\unit{ppb}$  in late April (day of the year (doy) $100-120$) and declines throughout the growing season until a minimum is reached in late summer/fall ($24-30\,\unit{ppb}$). The turning points lie approximately in June and November. The build-up of the ozone spring peak takes about $7.5\,\unit{months}$ from minimum to maximum. The most pronounced increase in \chem{\left<[O_3]\right>} coincident with rising temperatures and global irradiance. The depletion of ozone from maximum to minimum takes only $4.5\,\unit{months}$. The minimum occurs at $25\,\unit{ppb}$ in the beginning of September ($240-250\,\unit{doy}$).

The \chem{[O_3]} density distribution and the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} for Svanvik display a similar pattern as the generalized northern Fennoscandia climatology (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b). The spring peak occurs slightly earlier ($100\,\unit{doy}$) and amounts to $40\,\unit{ppb}$ while the annual average \chem{\left<[O_3]\right>} is $6.6\,\unit{ppb}$ lower than the Fennoscandic climatology. This is in line with a general decrease in tropospheric ozone with decreasing altitude. The earlier decline in \chem{\left<[O_3]\right>} may point to an influence of coniferous vegetation on the removal of ozone since \chem{CO_2} uptake by coniferous trees has been observed as early as \unit{doy} 100 \citep{TB:Kolari2007, TP:Wallin2013} compared to an estimated average begin of growing season ($\left<G_\mathrm{start}\right>$) on $(150\pm 14)\,\unit{doy}$ at Svanvik (see Section~\ref{subsec:do3se_parameters} for details). In July--September, ozone is occasionally almost completely depleted during nighttime. This hint to ozone uptake by vegetation even at low light intensities during midnight sun conditions in combination with stable planetary boundary layer conditions preventing mixing of ozone rich air. The derived \chem{\left<[O_3]\right>} is likely biased towards lower \chem{[O_3]} due to an increase in northern hemispheric tropospheric background \chem{[O_3]} since the 1990s. To account for this, we derive a bias correction factor by cross-calibrating with the time series taken at Esrange and Pallas (Section~\ref{subsec:anomalies}).

@@

\subsection{2018/19 anomalies}
\label{subsec:anomalies}
Using these established climatologies, we look at the anomalies in 2018/19 and discuss the differences between the two years. As weather extremes like in 2018 are likely to become more frequent with climate change, this will help to improve our understanding of a possible future increase in vulnerability of subarctic vegetation to ozone.

\subsubsection{Temperature, precipitation, and global irradiance}
\label{subsubsec:anomal_tpq}
In Fig.~\ref{fig:anomalies_svanvik}a), temperature anomalies at Svanvik are shown as a percentage of days warmer or colder than $\pm 1\,\sigma$ from the climatological mean for each month. Negative deviations from climatology are displayed as a negative percentage. The full annual positive/negative deviations are indicated in the respective corners (right upper/right lower). The hatched area between the dashed lines indicates the expected percentage of values falling above/below $\pm 1\,\sigma$ if a normal distribution is assumed ($15.9\,\unit{\%}$). The summer of 2018 was significantly warmer than average. Especially in May and July more than $40\,\unit{\%}$ of the days were significantly warmer. Significantly warmer days continued to occur in all months (August--November) except for October. March 2018 had many unusually cold days. The summer of 2019 was fairly average, however, significantly colder days in July occurred, while August had both significantly colder and warmer days. April 2019 had significantly warmer than average days.

@@

Similarly, the precipitation anomalies at Svanvik are displayed in Fig.~\ref{fig:anomalies_svanvik}b). Due to large interannual variability in observed precipitation, only a few days deviated significantly from the climatology on the $1 \sigma$ level. Therefore, we lowered the $\sigma$ level constraint. The percentage of days wetter/drier than $\pm \frac{1}{2}\,\sigma$ from the climatological mean are shown for each month. Negative deviations from the climatology are shown as a negative percentage and annual positive/negative deviations are indicated in the respective corners. The hatched area between the dashed lines indicates the expected percentage of values falling above $\pm\frac{1}{2}\,\sigma$ if a normal distribution is assumed ($30.8\,\unit{\%}$). Unlike the temperature anomalies, the picture is not as clear. While March 2018 had a significant number of days wetter than unusual, summer and fall (May/July, September/October) had a significant number of days drier than average. Overall, 2019 was rather average throughout the year, but with a significant number of days drier than normal (September/October) and a wetter December.

Similarly, the global irradiance anomalies at Svanvik are presented in Fig.~\ref{fig:anomalies_svanvik}c). The percentage of days above or below $\pm 1\,\sigma$ from the climatological mean is displayed for each month. The hatched area between the dashed lines indicates the expected percentage of values falling above/below $\pm 1\,\sigma$ if a normal distribution is assumed. In summer 2018 (May/July), global irradiance showed a significant number of days with higher than average irradiance, while 2019 was generally an average year with a significant number of days darker in spring (April/May).

@@

\subsubsection{Ozone}
\label{subsubsec:anomal_ozone}
To assess the significance of 2018/19 anomalies, we estimate the bias due to the change in the tropospheric background \chem{[O_3]} since the 1990s and account for the missing data in June 2018. To reconstruct these data, we performed a Reynolds decomposition (Eq.~(\ref{eq:reynolds_decomp})) and cross-calibrate with ozone observations at Esrange and Pallas (companion paper). In the following, we will describe the bias-correction.

In 2018, the probability density function of ozone concentration anomalies (\chem{\Delta[O_3]}) for Esrange and Pallas deviate clearly, but not significantly from the climatological mean (Supplement~Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_residuals}). On average, \chem{[O_3]} had been elevated by $(1.9\pm 5.7)\,\unit{ppb}$ and $(1.9\pm 5.4)\,\unit{ppb}$, respectively. Both distributions are slightly skewed towards positive deviations with $\frac{1}{4}$ of days (3. quantile, q3) falling above $5.8\,\unit{ppb}$ and $5.6\,\unit{ppb}$, respectively.

Similarly, we find a skew towards higher \chem{[O_3]} for Svanvik in 2018, while 2019 follows a normal distribution rather well (Supplement~Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_residuals-Svanvik}). In conjunction with the meteorological conditions, we therefore can assume that 2019 had been a normal year and that the annual average anomaly in 2019 can be used for the bias correction of the Svanvik climatology (\chem{\left<\Delta[O_3]\right>}~=~$(1.2\pm 5.0)\,\unit{ppb}$). To further strengthen this, we look at trends in ozone observations at Esrange and Pallas and compute the average difference between recent years and the mid 1990s (\chem{\Delta_{present-past}\left<[O_3]\right>}). Present is defined as the average over 2010--2018 and past as the time between the respective beginning of data taking until 1997. We do not find a clear trend. For Esrange \chem{\Delta_{present-past}\left<[O_3]\right>} amounts to $(1.5\pm 3.1)\,\unit{ppb}$ and for Pallas to $(-2.5\pm 3.7)\,\unit{ppb}$ \citep[][for comparison]{NILU2003, BER:Ruoho-Airola2015}. The observed trend at Esrange is compatible with our assumed bias correction for the Svanvik ozone climatology.

@@

In accordance to the above findings, we adjust the derived ozone climatology for Svanvik by adding a constant bias correction factor $\Delta\chem{[O_3]_\mathrm{bias-corr}} = 1.2\,\unit{ppb}$ and assess the monthly significance of the ozone concentration anomalies in 2018/19 (Fig.~\ref{fig:anomalies_svanvik}d)). The resulting percentage of days above/below $\pm 1\,\sigma$ from the bias-corrected climatological mean is displayed for each month in Fig~\ref{fig:data_svanvik_2018_2019}. Negative deviations from climatology are shown as a negative percentage. Note that missing bars for certain months are both due to data availability and anomalies not meeting the $1\sigma$ criteria. The the star indicates resulting anomalies after the reconstruction of missing data in July. In April/May 2018, ozone was significantly enhanced, while no significant enhancement is found in 2019. In both years, ozone concentrations were significantly lower than average in November. These results mean, that the extensive forest fires in Sweden did not lead to a significant enhancement in \chem{[O_3]} at Svanhovd in 2018. However, there are indications that the sites further south at Esrange and Pallas experienced more days with elevated ozone in 2018 than Svanvik.

>>>>

According to a report by the Norwegian Meteorological institute \citep{MetNOR2019}, the summer of 2018 was the warmest and driest ever recorded in eastern, western, and southern Norway. In the north (including Finnmark), it was amongst the warmest on record -- favorable conditions for ground-level ozone formation.
An unusually weak and northward shifted jet stream allowed for a persistent high-pressure system above northern Europe, including Fennoscandia which blocked the low-pressure systems for several consecutive months. In the period May -- July, southern Norway had temperatures $4\,\unit{^\circ C}$ above normal. While southern Norway had only about $60\,\unit{\%}$ of normal precipitation, northern Norway as a whole had close to normal precipitation, but with local variations.
Thermal stress on vegetation was exceptional not only in large parts of Fennoscandia but also in much of Europe, where the influence of the high-pressure system extended even over five months (April/May, July--September). These conditions gave rise to massive forest fires in different parts of Europe and thus an increase in ozone precursors. Boreal wildfires emit in addition to \chem{CO_2} also \chem{CO} ($\chem{[CO]}/\chem{[CO_2]} \propto 6-13\,\unit{\%}$) and VOCs ($\chem{[VOCs]}/\chem{[CO_2]} \propto 0.5-1.5\,\unit{\%}$) \citep{AE:Cofer1990}.
A total of 2079 forest fires of different sizes were registered in Norway in 2018, twice as many as in the preceding years 2016/17 \citep{DSB2019}, last accessed April 2020). In Sweden, about 500 fires had been reported (five times more than in a usual summer), and an estimated total of $25000\,\unit{hectare}$ burned down in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) \citep{SOU2019}. Coincident peak \chem{[O_3]} is found in ozone monitoring data from Svanhovd (Fig.~\ref{fig:data_svanvik_2018_2019} but also over whole northern Fennoscandia in July.
During the 2003 drought period, elevated ozone in Europe was promoted by a combination of various factors such as wildfires, reduced cloud cover (increased solar radiation), reduced dry deposition and turbulent mixing due to the stagnant weather conditions, and increased BVOC emissions \citep{JGR:Solberg2018}. It is likely that these factors also contributed to enhanced \chem{[O_3]} in 2018.\\

@@

The observation site at NIBIO Environment Centre Svanhovd and all related data are presented in Section~\ref{sec:data}. In Section~\ref{sec:stats}, we derive multi-annual means for ground-level ozone concentration and key environmental variables (temperature, precipitation, and global irradiance) based on in situ observations and evaluate the deviation of 2018/19 from the norm. We use the $\mathrm{DO_3SE}$ model to estimate the ozone uptake by natural and semi-natural vegetation to investigate the main environmental drivers for projected ozone damage risk in 2018 (Section~\ref{sec:do3se}). To this end, we derive bespoke parameterizations for subarctic species to assess systematic uncertainties arising from the use of generic parameterizations established from less climatically extreme conditions. In Section~\ref{sec:conc}, we will summarize our results and discuss how this research can support improvements in future assessments of pollution risk and more generally biogeochemical cycling.

@@

Hourly averaged $2\,\unit{m}$ temperatures above $20\,\unit{^\circ C}$ occurred more regularly in July 2018 than in 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}b). In 2018, spring temperature regularly rose above freezing only in May, while in 2019 this occurred already early in March/April.
More rain events with accumulated daily precipitation ($\sum_d \mathrm{Precip}$) above $10\,\unit{mm}$ occurred in the summer of 2018 compared to 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}c).
Qualitatively, global irradiance ($Q_0$) displayed in Fig.~\ref{fig:data_svanvik_2018_2019}d) was higher in May and July 2018 compared to 2019, while June 2019 showed higher irradiance than 2018. A high global irradiance in most cases is the result of a low cloud fraction. In both years, the maximum recorded $Q_0$ was $750\,\unit{W\,m^{-2}}$ and reached in June.

@@

A comprehensive overview over all past and present ozone observation sites in northern Fennoscandia is given in Table~\ref{tab:ebas_obs} and Fig.~\ref{fig:station_map_fennoscandia}, respectively. All stations with available long-term observations are located at higher altitude than Svanhovd. In the absence of local precursors, the stratification of the atmosphere usually leads to a higher abundances in ozone at higher altitude \citep[e.g.,][]{AB:Klingberg2009}. The periphery vegetation at all sites is similar and consists mainly of pine forests, birch, perennial grassland, and heath. However, on closer inspection the vegetation differs substantially. Ozone measurements at Pallas are taken on top of mount Sammaltunturi, $100\,\unit{m}$ above treeline (pine forest), and the surrounding vegetation mainly consists of low vascular plants, mosses, and lichen \citep{BER:Hatakka2003}. At Svanhovd, the ozone measurements are taken amidst a meadow surrounded by agriculturally managed herbaceous plants and grass adhering to the \citet{WMOGuide2018} standards (Fig.\ref{fig:svanhovd_research_station}). In conjunction with the location at lower altitude, this results in generally lower \chem{[O_3]} at Svanhovd compared to Pallas.

Data from before the 1990s are available solely from Svanvik and Jergul. These data, however, do not follow the high-quality standard procedures implemented nowadays and have to be treated with care \citep{NILU2003}. Especially, ozone monitors did not undergo regular re-calibration. This probably leads to drifts in the observed data and may impose a false trend. In 1997, the ozone monitoring at Jergul was relocated downstream of the river Karasjokka to a site close by Karasjok. Although the same equipment has been used, this might have introduced a baseline shift in the combined data series, denoted as Jergul/Karasjok, which would be particularly problematic if these were the sole data for comparison. Only a few data have been taken upstream the Pasvik river at Janiskovski, hence these will not be taken into further consideration, but are listed for completeness. There is an overlap of about $4\,\unit{years}$ between data taken at Esrange, Jergul/Karasjok, and Svanvik (Supplement~Fig.~\ref{fig:ozone_timesseries_fenoscandic_obs}).

@@

\subsubsection{Temperature, precipitation, and global irradiance}
\label{subsubsec:clim_temp_prec}
The location of Svanvik at $69.45\,\unit{^\circ N}$ suggests a subarctic climate. The derived climate diagram based on climatological data (1992--2012) (Fig.~\ref{fig:climatediagram}) supports this very well. Monthly averaged $2\,\unit{m}$ temperatures (red line) are displayed with a standard error of the mean (SE) as error band. We chose SE in this case due to the large interannual variability in temperature reflected in the standard deviation. Averaged monthly accumulated precipitation (blue bars) is shown with standard deviation as error bars. Months with average temperatures below freezing are denoted with a star. As can be seen, temperatures stay below freezing for 5 consecutive months, while only 2 months breach $10\,\unit{^\circ C}$ regularly (July, August), satisfying the conditions for K\"{o}ppen's climate classification of a regular subarctic climate (Dfc) \citep[e.g.][]{SD:Beck2018}. The highest monthly average temperature is $(13.1\pm 1.1)\,\unit{^\circ C}$ in July and the lowest $(-12.8\pm 2.0)\,\unit{^\circ C}$ in February. In the period 1992--2012, the coldest measured temperature was $-45.2\,\unit{^\circ C}$ observed January 27, 1999, while the highest temperature ($29.4\,\unit{^\circ C}$) occurred July 16 the same year.

@@

The derived climatology for global irradiance ($Q_0$) at Svanvik (Appendix Fig.~\ref{fig:global_rad_clim}a)) approaches zero in December/January (polar night) and reaches its maximum in June/July (midnight sun conditions). The highest observed averaged daily maximum ($\left<Q_0^\mathrm{max}\right>$) amounts to $800\,\unit{W\,m^{-2}}$, while the maximum of the averaged daily mean ($\left<Q_0^\mathrm{mean}\right>$) is $450\,\unit{W\,m^{-2}}$ (regularly reached in May--July). The maximum of the averaged daily minimum ($\left<Q_0^\mathrm{min}\right>$) lies at about $200\,\unit{W\,m^{-2}}$.

\subsubsection{Ozone}
\label{subsubsec:clim_ozone}
In the following, we derive an ozone climatology (\chem{\left<[O_3]\right>}) for Svanvik and for northern Fennoscandia. To this end, we first calculate a pairwise correlation between the different stations. The correlation of \chem{[O_3]} between Esrange, Pallas, and Jergul/Karasjok is fairly high ($r^2\approx 0.8$, Appendix Fig.~\ref{fig:density_distribution}). Therefore, we select these data to compute the generalized ozone climatology for northern Fennoscandia. The correlation between Svanvik with any of the active monitoring sites in 2018 (Esrange and Pallas) is rather low, 0.4 and 0.6, respectively. Thus, conditions at Svanvik cannot directly be represented by measurements taken at either Esrange or Pallas.

In Figure~\ref{fig:ozone_climatology_fenoscandic_obs_spline}, two dimensional density distributions of \chem{[O_3]} for the combined data (Esrange, Jergul/Karasjok, Pallas) (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}a) and Svanvik (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) are shown. Darker colors indicate higher probability to observe these values. On top of the density distributions, a $10\,\unit{days}$ average of daily mean (\chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}}) is displayed together with $1 \sigma$ uncertainties and SE, respectively. Splines have been fitted through the data for easier inter-comparison.

The generalized northern Fennoscandic ozone climatology peaks at $48\,\unit{ppb}$  in late April (day of the year (doy) $100-120$) and declines throughout the growing season until a minimum is reached in the beginning of September ($240-250\,\unit{doy}$) and amounts to $24-30\,\unit{ppb}$. The turning points lie approximately in June and November. The build-up of the ozone spring peak takes about $7.5\,\unit{months}$ from minimum to maximum. The most pronounced increase in \chem{\left<[O_3]\right>} coincides with rising temperatures and global irradiance. The depletion of ozone from maximum to minimum takes only $4.5\,\unit{months}$.

The \chem{[O_3]} density distribution and the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} for Svanvik display a similar pattern as the generalized northern Fennoscandia climatology (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b). The spring peak occurs slightly earlier ($100\,\unit{doy}$) and amounts to $40\,\unit{ppb}$ while the annual average \chem{\left<[O_3]\right>} is $6.6\,\unit{ppb}$ lower than the Fennoscandic climatology. This is in line with a general decrease in tropospheric ozone with decreasing altitude. The earlier decline in \chem{\left<[O_3]\right>} may point to an influence of coniferous vegetation on the removal of ozone since \chem{CO_2} uptake by coniferous trees has been observed as early as \unit{doy} 100 \citep{TB:Kolari2007, TP:Wallin2013}. The estimated average begin of growing season for deciduous trees ($\left<G_\mathrm{start}\right>_\mathrm{decid}$) is $(150\pm 14)\,\unit{doy}$ at Svanvik (see Section~\ref{subsec:do3se_parameters} for details). In July--September, ozone is occasionally almost completely depleted during nighttime. This hints to ozone uptake by vegetation even at low light intensities during midnight sun conditions in combination with stable planetary boundary layer conditions preventing mixing of ozone rich air. The derived \chem{\left<[O_3]\right>} is likely biased towards lower \chem{[O_3]} due to an increase in northern hemispheric tropospheric background \chem{[O_3]} since the 1990s. To account for this, we derive a bias correction factor by cross-calibrating with the time series taken at Esrange and Pallas (Section~\ref{subsec:anomalies}).

@@

\subsection{2018/19 anomalies}
\label{subsec:anomalies}
In the following, we look at the 2018/19 anomalies and discuss the differences between the two years. We also conduct a bias correction for the Svanvik \chem{\left<[O_3]\right>.} As weather extremes like in 2018 are likely to become more frequent with climate change, this will help to assess possible future changes in the vulnerability of subarctic vegetation to ozone.

\subsubsection{Temperature, precipitation, and global irradiance}
\label{subsubsec:anomal_tpq}
In Fig.~\ref{fig:anomalies_svanvik}a), temperature anomalies for Svanvik are shown as a percentage of days warmer or colder than $\pm 1\,\sigma$ from the climatological mean for each month. Negative deviations from climatology are displayed as a negative percentage. The annual positive/negative deviations are indicated in the respective corners (right upper/right lower). The hatched area between the dashed lines indicates the expected percentage of values falling above/below $\pm 1\,\sigma$ if a normal distribution is assumed ($15.9\,\unit{\%}$). We find, that the summer of 2018 was significantly warmer than average. Especially in May and July more than $40\,\unit{\%}$ of the days were significantly warmer. Significantly warmer days continued to occur in all months (August--November) except for October. March 2018 had many unusually cold days. The summer of 2019 was fairly average, however, significantly colder days occurred in July, while August had both significantly colder and warmer days. April 2019 had significantly warmer than average days.

@@

Similarly, the precipitation anomalies for Svanvik are displayed in Fig.~\ref{fig:anomalies_svanvik}b). Due to the large interannual variability in observed precipitation, only a few days deviated significantly from the climatology on the $1 \sigma$ level. Therefore, we lowered the $\sigma$ level constraint to $\pm \frac{1}{2}\,\sigma$. The percentage of days wetter/drier than $\pm \frac{1}{2}\,\sigma$ are shown for each month. Positive deviation refers to days wetter than average and negative deviation to drier than average. The hatched area between the dashed lines indicates the expected percentage of values falling above/below $\pm\frac{1}{2}\,\sigma$ if a normal distribution is assumed ($30.8\,\unit{\%}$). Unlike the temperature anomalies, the picture is not as clear. While March 2018 had a significant number of days wetter than unusual, summer and fall (May/July, September/October) had a significant number of days drier than average. 2019 was rather average throughout the year, but with a significant number of days drier than normal in September/October and wetter in December.

The global irradiance anomalies for Svanvik are presented in Fig.~\ref{fig:anomalies_svanvik}c). The percentage of days above or below $\pm 1\,\sigma$ from the climatological mean is displayed for each month. The hatched area between the dashed lines indicates the expected percentage of values falling above/below $\pm 1\,\sigma$ if a normal distribution is assumed. In summer 2018 (May/July), global irradiance showed a significant number of days with higher than average irradiance, while 2019 was generally an average year with a significant number of days darker in spring (April/May).

@@

\subsubsection{Ozone}
\label{subsubsec:anomal_ozone}
To assess the significance of 2018/19 anomalies, we estimate the bias due to the change in the tropospheric background \chem{[O_3]} since the 1990s and account for the missing data in June 2018. To reconstruct these missing data, we performed a Reynolds decomposition (Eq.~(\ref{eq:reynolds_decomp})) and cross-calibrate with ozone observations at Esrange and Pallas (companion paper). In the following, we will describe the bias-correction and discuss the significance of \chem{[O_3]} enhancements in 2018.

In 2018, the probability density functions (PDFs) of ozone concentration anomalies (\chem{\Delta[O_3]}) for Esrange and Pallas deviate clearly, but not significantly from the climatological mean (Supplement~Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_residuals}). On average, \chem{[O_3]} had been elevated by $(1.9\pm 5.7)\,\unit{ppb}$ and $(1.9\pm 5.4)\,\unit{ppb}$, respectively. Both distributions are slightly skewed towards positive deviations with $\frac{1}{4}$ of days (3. quantile, q3) falling above $5.8\,\unit{ppb}$ and $5.6\,\unit{ppb}$, respectively.

We find a similar skew towards higher \chem{[O_3]} for Svanvik in 2018, while 2019 follows a normal distribution rather well (Supplement~Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_residuals-Svanvik}). In conjunction with the meteorological conditions, we therefore assume that 2019 had been a normal year and the annual average anomaly in 2019 (\chem{\left<\Delta[O_3]\right>}~=~$(1.2\pm 5.0)\,\unit{ppb}$) can be used for the bias correction. We also consult the trends in ozone observations at Esrange and Pallas to strengthen this assumption and compute the average difference between recent years and the mid 1990s (\chem{\Delta_{present-past}\left<[O_3]\right>}). Herein, present is defined as the average over 2010--2018 and past as the time between the respective beginning of data taking until 1997. We do not find a clear trend. For Esrange \chem{\Delta_{present-past}\left<[O_3]\right>} amounts to $(1.5\pm 3.1)\,\unit{ppb}$ and for Pallas to $(-2.5\pm 3.7)\,\unit{ppb}$ \citep[][for comparison]{NILU2003, BER:Ruoho-Airola2015}. Only the observed trend at Esrange is compatible with our assumed bias correction for the Svanvik ozone climatology.

@@

In accordance to the above findings, we adjust the derived ozone climatology for Svanvik by adding a constant bias correction factor $\Delta\chem{[O_3]_\mathrm{bias-corr}} = 1.2\,\unit{ppb}$ and assess the monthly significance of the ozone concentration anomalies in 2018/19. The resulting percentage of days above/below $\pm 1\,\sigma$ from the bias-corrected climatological mean is displayed for each month in Fig.~\ref{fig:anomalies_svanvik}d). Negative deviations from climatology are shown as a negative percentage. Note that missing bars for certain months are both due to data availability and anomalies not meeting the $1\sigma$ criteria. The star indicates results after the reconstruction of missing data in July. In April/May 2018, ozone was significantly enhanced, while no significant enhancement is found in 2019. In both years, ozone concentrations were significantly lower than average in November. These results indicate that the extensive forest fires in Sweden did not contribute to a significant enhancement in \chem{[O_3]} at Svanhovd in 2018, but the stagnant meteorological conditions in April/May in combination with a late start of the growing season. However, there are indications that the sites further south at Esrange and Pallas experienced more days of elevated ozone in 2018 than Svanvik.
